---
title: "Identification and classification of duplicated genes"
author: 
  - name: Fabricio Almeida-Silva
    affiliation: 
    - VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
    - Department of Plant Biotechnology and Bioinformatics, Ghent University, Ghent, Belgium
  - name: Yves Van de Peer
    affiliation: 
    - VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
    - Department of Plant Biotechnology and Bioinformatics, Ghent University, Ghent, Belgium
    - College of Horticulture, Academy for Advanced Interdisciplinary Studies, Nanjing Agricultural University, Nanjing, China
    - Center for Microbial Ecology and Genomics, Department of Biochemistry, Genetics and Microbiology, University of Pretoria, Pretoria, South Africa
output: 
  BiocStyle::html_document:
    toc: true
    number_sections: yes
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Identification and classification of duplicated genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Introduction

Intro.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
  }

BiocManager::install("doubletrouble")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

Then, you can load the package:

```{r load_package}
library(doubletrouble)
```

# Data description

In this vignette, we will use protein sequences (primary transcripts only)
and genome annotation for the yeast species *Saccharomyces cerevisiae*
and *Candida glabrata*. Data were obtained from 
Ensembl Fungi release 54 [@yates2022ensembl].

The example data sets are stored in the following objects:

- **yeast_seq:** A list of `AAStringSet` objects with elements
named *Scerevisiae* and *Cglabrata*.

```{r data1}
# Load list of DIAMOND tabular output
data(yeast_seq)
head(yeast_seq)
```

- **yeast_annot:** A `GRangesList` object with elements 
named *Scerevisiae* and *Cglabrata*.

```{r ex_data}
# Load annotation list processed with syntenet::process_input()
data(yeast_annot)
head(yeast_annot)
```

Our goal here is to identify and classify duplicated genes in 
the *S. cerevisiae* genome. The *C. glabrata* genome will be used as an outgroup
to identify transposed duplicates later in this vignette.

# Data preparation

First of all, we need to process the list of protein sequences and gene ranges
to detect synteny with `r BiocStyle::Biocpkg("syntenet")`. We will do that
using the function `process_input()` from 
the `r BiocStyle::Biocpkg("syntenet")` package.

```{r process_input}
library(syntenet)

# Process input data
pdata <- process_input(yeast_seq, yeast_annot)

# Inspect the output
names(pdata)
pdata$seq
pdata$annotation
```

The processed data is represented as a list with the elements `seq` and
`annotation`, each containing the protein sequences and gene ranges for
each species, respectively.

Finally, we need to perform pairwise sequence similarity searches to
identify the whole set of paralogous gene pairs. We can do this 
using the function `run_diamond()` from the `r BiocStyle::Biocpkg("syntenet")`
package [^1], setting `compare = "intraspecies"` to perform only intraspecies
comparisons.

[^1]: **Note:** you need to have DIAMOND installed in your machine to run
this function. If you don't have it, you can use 
the `r BiocStyle::Biocpkg("Herper")` package to install DIAMOND in a Conda
environment and run DIAMOND from this virtual environment.

```{r run_diamond_intraspecies}
data(diamond_intra)

# Run DIAMOND in sensitive mode for S. cerevisiae only
if(diamond_is_installed()) {
    diamond_intra <- run_diamond(
        seq = pdata$seq["Scerevisiae"],
        compare = "intraspecies", 
        outdir = file.path(tempdir(), "diamond_intra"),
        ... = "--sensitive"
    )
}

# Inspect output
names(diamond_intra)
head(diamond_intra$Scerevisiae_Scerevisiae)
```

And voilà! Now that we have the DIAMOND output and the processed annotation,
you can classify the duplicated genes.

# Classifying duplicated gene pairs and genes

To classify duplicated gene pairs based on their modes of duplication,
you will use the function `classify_gene_pairs()`. As input, you need to pass:

- The processed annotation list as returned by
`syntenet::process_input()`;
- A list of data frames with DIAMOND (or BLASTp, etc.) tabular output
as returned by `syntenet::run_diamond()`.


The function `classify_gene_pairs` incorporates 3 classification schemes:

1. **Binary:** Duplicated gene pairs are classified as either derived
from whole-genome duplications (WGD) or small-scale duplications (SSD).
2. **Expanded:** SSD-derived genes are divided into 3 subcategories:
tandem duplication (TD), proximal duplication (PD), and 
dispersed duplication (DD).
3. **Full:** Similar to the **expanded** scheme, but an additional subcategory
is used for SSD-derived gene pairs: transposed duplication (TRD). [^2] 

[^2]: **Note:** To identify duplicated genes derived from transpositions, 
you will need to input the DIAMOND output of a similarity search between your 
species of interest and an outgroup. See Section \@ref(full) for details.

## Binary classification

To classify duplicated gene pairs into WGD- or SSD-derived, use `binary = TRUE`
in `classify_gene_pairs()`.

```{r binary_classification}
# Binary classification
c_binary <- classify_gene_pairs(
    blast_list = diamond_intra,
    annotation = pdata$annotation,
    binary = TRUE
)

# Inspecting the output
names(c_binary)
head(c_binary$Scerevisiae)

# How many pairs are there for each duplication mode?
table(c_binary$Scerevisiae$type)
```

The function returns a list of data frames, each containing the duplicated
gene pairs and their modes of duplication for each species.

## Expanded classification (SSD &rarr; TD, PD, DD)

To subdivide SSD-derived gene pairs into TD-, PD-, and DD-derived, 
use `binary = FALSE` in `classify_gene_pairs()`.

```{r expanded_classification}
# Expanded classification
c_expanded <- classify_gene_pairs(
    blast_list = diamond_intra,
    annotation = pdata$annotation,
    binary = FALSE
)

# Inspecting the output
names(c_expanded)
head(c_expanded$Scerevisiae)

# How many pairs are there for each duplication mode?
table(c_expanded$Scerevisiae$type)
```

## Full classification (SSD &rarr; TD, PD, TRD, DD) {#full}

To find transposon-derived duplicates, first you will need to perform a 
similarity search of your target species against an outgroup species.
You can do this with `syntenet::run_diamond()`. For the parameter `compare`,
you will pass a 2-column data frame specifying the comparisons to be made. [^3]

[^3]: **Pro tip:** If you want to identify and classify duplicated genes for
multiple species in batch, you must include the outgroup for each of them
in the comparisons data frame.

Here, we will identify duplicated gene pairs for *Saccharomyces cerevisiae*
using *Candida glabrata* as an outgroup.

```{r blast_interspecies}
data(diamond_inter) # load pre-computed output in case DIAMOND is not installed

# Create data frame of comparisons to be made
comparisons <- data.frame(
    species = "Scerevisiae",
    outgroup = "Cglabrata"
)
comparisons

# Run DIAMOND for the comparison we specified
if(diamond_is_installed()) {
    diamond_inter <- run_diamond(
        seq = pdata$seq,
        compare = comparisons,
        outdir = file.path(tempdir(), "diamond_inter"),
        ... = "--sensitive"
    )
}

names(diamond_inter)
head(diamond_inter$Scerevisiae_Cglabrata)
```

Now, we will pass this interspecies DIAMOND output as an argument to 
the parameter `blast_inter` of `classify_gene_pairs()`.

```{r full_classification}
# Full classification
c_full <- classify_gene_pairs(
    blast_list = diamond_intra,
    annotation = pdata$annotation,
    binary = FALSE,
    blast_inter = diamond_inter
)

# Inspecting the output
names(c_full)
head(c_full$Scerevisiae)

# How many pairs are there for each duplication mode?
table(c_full$Scerevisiae$type)
```

# Classifying genes into unique modes of duplication

If you look carefully at the output of `classify_gene_pairs()`, you will notice
that some genes appear in more than one duplicate pair, and these pairs can
have different duplication modes assigned. There's nothing wrong with it.
Consider, for example, a gene that was originated by a whole-genome duplication
event some 60 million years ago, and then it underwent a tandem duplication
5 million years ago. In the output of `classify_gene_pairs()`, you'd see
this gene in two pairs, one with **WGD** in the *type* column, and one
with **TD**.

If you want to assign each gene to a unique mode of duplication, you can
use the function `classify_genes()`. This function assigns duplication modes
hierarchically in this level of priority: WGD > TD > PD > TRD > DD.

The input for `classify_genes()` is the list of gene pairs returned by
`classify_gene_pairs()`.

```{r classify_genes}
# Classify genes into unique modes of duplication
c_genes <- classify_genes(c_full)

# Inspecting the output
names(c_genes)
head(c_genes$Scerevisiae)

# Number of genes per mode
table(c_genes$Scerevisiae$type)
```

# Calculating Ka, Ks, and Ka/Ks for duplicated gene pairs

You can use the function `pairs2kaks()` to calculate rates of nonsynonymous 
substitutions per nonsynonymous site (Ka), synonymouys substitutions per
synonymous site (Ks), and their ratios (Ka/Ks). These rates are calculated
using the Bioconductor package `r BiocStyle::Biocpkg("MSA2dist")`, which
implements all codon models in KaKs_Calculator 2.0 [@wang2010kaks_calculator].


For the purpose of demonstration, we will only calculate Ka, Ks, and Ka/Ks
for 5 WGD-derived gene pairs. The CDS for WGD-derived genes were obtained
from Ensembl Fungi [@yates2022ensembl], and 
they are stored in `cds_scerevisiae`.

```{r kaks_calculation}
data(cds_scerevisiae)
head(cds_scerevisiae)

# Store DNAStringSet object in a list
cds_list <- list(Scerevisiae = cds_scerevisiae)

# Keep only top give gene pairs for demonstration purposes
gene_pairs <- list(Scerevisiae = c_binary$Scerevisiae[1:5, ])

# Calculate Ka, Ks, and Ka/Ks
kaks <- pairs2kaks(gene_pairs, cds_list)

# Inpsect the output
head(kaks)
```


# Identifying and visualizing Ks peaks

Peaks in Ks distributions typically indicate whole-genome duplication (WGD) 
events, and they can identified by fitting Gaussian mixture models (GMMs) to 
Ks distributions. In `r BiocStyle::Githubpkg("doubletrouble")`, this can be 
performed with the function `find_ks_peaks()`.


However, because of saturation at higher Ks values, only **recent WGD**
events can be reliably identified from Ks 
distributions [@vanneste2013inference]. Recent WGD events are commonly found 
in plant species, such as maize, soybean, apple, etc.
Although the genomes of yeast species have signatures of WGD,
these events are ancient, so it is very hard to find evidence for them
using Ks distributions. [^4] 

[^4]: **Tip:** You might be asking yourself: "How does one identify ancient
WGD, then?". A common approach is to look for syntenic blocks (i.e.,
regions with conserved gene content and order) within genomes. This is what
`classify_gene_pairs()` does under the hood to find WGD-derived gene pairs.

To demonstrate how you can find peaks in Ks distributions
with `find_ks_peaks()`, we will use a data frame containing Ks values for
duplicate pairs in the soybean (*Glycine max*) genome, which has undergone 
2 WGDs events ~13 and ~58 million years ago [@schmutz2010genome]. 
Then, we will visualize Ks distributions with peaks using the function
`plot_ks_peaks()`.

First of all, let's look at the data and have a quick look at the distribution.

```{r ks_eda}
# Load data and inspect it
data(gmax_ks)
head(gmax_ks)

# Plot distribution
library(ggplot2)
ggplot(gmax_ks, aes(x = Ks)) +
    geom_histogram(color = "black", fill = "grey90", bins = 50) +
    theme_bw()
```

By visual inspection, we can see 2 or 3 peaks. Based on our prior knowledge,
we know that 2 WGD events have occurred in the ancestral of the *Glycine* genus
and in the ancestral of all Fabaceae, which seem to correspond to the
peaks we see at Ks values around 0.1 and 0.5, respectively. There could be
a third, flattened peak at around 1.6, which would represent the WGD shared
by all eudicots. Let's test which number of peaks has more support: 2 or 3.

```{r find_ks_peaks}
# Find 2 and 3 peaks and test which one has more support
peaks <- find_ks_peaks(gmax_ks$Ks, npeaks = 2:3, verbose = TRUE)
names(peaks)
str(peaks)

# Visualize Ks distribution
plot_ks_peaks(peaks)
```

As we can see, the presence of 3 peaks is more supported (lowest BIC). The
function returns the mean, variance and amplitude of mixture components 
(i.e., peaks), as well as the Ks distribution itself.

Now, suppose you just want to get the first 2 peaks. You can do that by
explictly saying to `find_ks_peaks()` how many peaks there are. You will
probably need to filter you Ks values by a maximum value as well.

```{r find_peaks_explicit}
# Find 2 peaks ignoring Ks values > 1
peaks <- find_ks_peaks(gmax_ks$Ks, npeaks = 2, max_ks = 1)
plot_ks_peaks(peaks)
```

**Important consideration on GMMs and Ks distributions:**
Peaks identified with GMMs should not be blindly regarded as "the truth".
Using GMMs to find peaks in Ks distributions can lead to problems such as
overfitting and overclustering [@tiley2018assessing]. Some general 
recommendations are:

1. Use your prior knowledge. If you know how many peaks there are (e.g.,
based on literature evidence), just tell the number to `find_ks_peaks()`.
GMMs can incorrectly identify more peaks than the actual number.

2. Test the significance of each peak with SiZer maps [@chaudhuri1999sizer].
This can be done with the function `SiZer()` from 
the R package `r BiocStyle::CRANpkg("feature")`.


# Session information {.unnumbered}

This document was created under the following conditions:

```{r session_info}
sessionInfo()
```

# References {.unnumbered}

